Episodic memory: mental time travel or a quantum `memory wave' function?

### 2020-09-18 paper/main.tex
working on word embedding text

\subsection*{Defining the geometries of ongoing thoughts and memories}
%JRM STOPPED HERE

>>>>

\subsection*{Defining the geometries of ongoing thoughts and memories}
Defining geometric spaces that can appropriately capture the conceptual contents of our thoughts and experiences has become a major focus of computational science.  An class of models that have shown particular promise over the past several decades is broadly referred to as \textit{word embedding models} (Fig.~\ref{fig:embedding-models}).  The fundamental goal of these models is to take as input a sequence of words (text) and produce as output one or more feature vectors whose coordinates reflect their conceptual (semantic) content.  For example, when two similar documents are passed through these models, the goal is to produce two feature vectors that are closer (e.g., in Euclidean distance) than the feature vectors for two dissimilar documents.

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{LandDuma97} attempted to compute these embeddings using word co-occurance statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

A related approach, termed \textit{Word Association Spaces}~\citep[WAS;][]{NelsEtal04, SteyEtal04} replaces the word count matrix with data from a large free-association experiment.  In the experiment, each participant was asked to free-associate (i.e., say the first word that came to mind) in response to a series of cue words.  The authors then constructed a number-of-cue-words by number-of-unique-response-words matrix whose entries reflected the proportion of times each cue word (row) produced a given response word (column) as an associate.  After using SVD to decompose this \textit{word frequency matrix}, they obtained low-dimensional embeddings of each cue word and response word.  In one sense, WAS provides a closer reflection of how people think about the meanings of words than LSA, since nearby WAS vectors reflect either cue words that produced similar responses, or response words that were evoked by similar cues.  In other words, WAS derives embeddings directly from people's semantic judgement behaviors.  However, because WAS is based on a large amount of experimental data, updating the embeddings requires collecting new data, which can be slow and costly.  As the target application or population diverges from the experimental population (e.g., over time, across cultures, demographics, etc.) the WAS embeddings may become less accurate or relevant.  By contrast, LSA may be re-fit relatively quickly to a new corpus, enabling the resulting embeddings to stay current.  Further, because LSA may be applied to any corpus, word embeddings may be easily computed and/or compared across languages, times, subject areas, etc.

\begin{figure}[tp]
\centering
\includegraphics[width=0.7\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A. Matrix factorization models.} Per-document (or per-participant) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflect the similarities between documents, and the columns of the word embeddings matrix reflect similarities between words.  \textbf{B. Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K \times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in the number-of-documents $\times K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C. Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements.  \textit{Transformer} models take a sequence as input and output a new modified sequence~\citep[diagram adapted from][]{ViswEtal17}.  The transformed outputs distort the original inputs in a way that often enables subseqeuent components of a deep text embedding model to compute more stable and meaningful embeddings.  \textit{Deep averaging networks} pool information from different observations in the original data in order to magnify small but meaningful differences.  \textit{Recurrent neural networks} (and variants) are often used to model temporal relations in sequence data~\citep[the deep averaging network and recurrent neural network diagrams are loosely inspired by figures from][]{IyyeEtal15}.  \textit{Deep autoencoders} attempt to reconstruct inputted patterns after passing them through a series of increasingly smaller layers in an encoding network.  The resulting activation patterns in the discriminator layer, which contains the fewest number of neurons, may be used as a low-dimensional embedding of the input.  The discriminator's activations are then passed through a decoder network which contains successively larger layers.  Finally, a reconstructed output is produced.  After fitting this model, outputs may be estimated from arbitrary activations of the discriminator layer~\citep[diagram loosly inspired by][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

% topic models-- deeper themes, stress interpretability.  example: bat (animal, sports equipment, verb).  "As the mutant vampire bat swooped low to attack, the explorer took a mighty swing of her trusty baseball bat.  The swing connected, stunning the creature before it scrambled about on the ground and flew away.  She didn't even bat an eye-- as the dark forces grew in power, such things had become increasingly common, nearly to the point of mundanity."

%deep learning approaches
